---
title: "Covariates"
description: >
  Specify participant-level predictors and understand how they enter the
  hierarchical kinetics model.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Covariates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

Covariates describe systematic differences between participant profiles. They
shift population kinetic parameters; participant random effects then describe
remaining heterogeneity around those conditional population curves.

```{r, echo=FALSE, out.width="98%", fig.cap="The default hierarchy. Biomarker-specific population kinetics are modified by participant covariates and selected non-centred participant effects, producing individual latent trajectories and finally measured or censored observations."}
doc_figure("documentation-hierarchy.svg")
```

## Formula and model matrix

Use an ordinary one-sided R formula:

```{r, collapse=FALSE}
library(epikinetics)
dat <- doc_delta_data()

prepared <- prepare_epikinetics_data(
  dat,
  formula = ~ infection_history,
  lower_limit = 5,
  upper_limit = 2560
)

prepared$mappings$reference_levels
prepared$mappings$design_columns
head(model.matrix(prepared))
```

Categorical variables use the active R contrasts, continuous variables remain
on their supplied scale, and supported interactions are encoded by
`model.matrix()`. Character predictors are converted to factors. Covariates
must be constant within participant.

Use a conventional intercept formula. `epikinetics` removes that design
intercept internally because each biomarker already has population intercept
parameters. A no-intercept formula such as
`~ 0 + infection_history` would be redundant and is rejected.

For continuous variables, centre or rescale in the input data when a one-unit
coefficient is not scientifically useful. Keeping this transformation explicit
makes both coefficients and future `newdata` easier to interpret.

## Meaning of a coefficient

For baseline, a coefficient is an additive shift in expected log2 response;
`2^coefficient` is the response-scale ratio. Positive timing and rate
quantities use a log link, so `exp(coefficient)` is their multiplicative
shift. This preserves positivity and the ordering of peak and switch times.

By default, a non-trivial formula modifies all six kinetic quantities, matching
the original scientific model. A narrower scientific hypothesis can name only
the affected quantities:

```{r, eval=FALSE}
prepare_epikinetics_data(
  dat,
  formula = ~ infection_history,
  covariate_parameters = c("baseline", "late_waning_rate")
)
```

Regression effects are shared across biomarkers in the current model.
`posterior_parameters(fit, level = "regression")` returns the numeric design
column, represented term/level, reference level, link-scale coefficient, and
interpretable transformed effect.

## Covariates in prediction

The prepared object retains its terms, assignments, contrasts, factor levels,
and labelled design columns. Prediction therefore applies exactly the encoding
used during fitting:

```{r}
prediction_grid(prepared)
```

The default grid keeps categorical combinations observed among participants and
fixes continuous predictors at participant-level medians. These are
*conditional* population profiles, not averages over the sample. Explicit
`newdata` can select fitted profiles; unknown factor levels fail early.
Categorical profile columns automatically determine facets in the default
population plot.

The default participant random effects apply to baseline, boost rate, early
waning, and late waning. Peak time and switch duration are shared across
participants after conditioning on covariates. This is an overridable modelling
assumption, described fully in
[Kinetics model and statistical structure](model.html).

Continue with [Population-level kinetics](population-kinetics.html) to see how
these profiles become trajectories.
