---
title: "A guide to biokinetics input and output data"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{A guide to biokinetics input and output data}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Input data

The model requires time series data about individual titre readings, along with last exposure times. Times can be relative (e.g. day of study) or absolute (i.e. precise calendar dates).
This is provided via the `data` argument when initialising an object of class [biokinetics](../reference/biokinetics.html), which must be a [data.table](https://rdatatable.gitlab.io/data.table/reference/data.table.html) containing the following columns:

| name         | type                 | description                                                                                | required |
|--------------|----------------------|--------------------------------------------------------------------------------------------| -------- |
| pid          | numeric or character | Unique identifier to identify a person across observations                                 | T|
| day          | integer or date      | The day of the observation. Can be a date or an integer representing a relative day of study | T|
| last_exp_day | integer or date      | The most recent day on which the person was exposed. Must be of the same type as the 'day' column | T|
| titre_type   | character            | Name of the titre or biomarker                                                             | T|
| value        | numeric              | Titre value                                                                                | T|
| censored     | -1, 0 or 1           | Optional column. Whether this observation should be treated as censored: -1 for lower, 1 for upper, 0 for none. | F|

It can also contain further columns for any covariates to be included in the model. The data files installed with this package have additional columns infection_history, last_vax_type, and exp_num.

The model also accepts a covariate formula to define the regression model. The variables in the formula must correspond to column names in the dataset. Note that all variables will be treated as **categorical variables**; that is, converted to factors regardless of their input type.

Note also that the `value` column is assumed to be on a natural scale by default, and will be converted to a log scale for model fitting. If your data
is already on a log scale, you must pass the `log=TRUE` argument when initialising the biokinetics class. See [biokinetics](../reference/biokinetics.html).

## Example

```{r}
dat <- data.table::fread(system.file("delta_full.rds", package = "epikinetics"))
head(dat)
```

# Output data

After fitting a model, a [CmdStanMCMC](https://mc-stan.org/cmdstanr/reference/CmdStanMCMC.html) object is returned. This means
that users who are already familiar with `cmdstanr` are free to do what they want with the fitted model.

**Important!**
**If you provide data on a natural scale, it will be converted to a base2 log scale before inference is performed. This means that if working
directly with the fitted `CmdStanMCMC` all values will be on this scale. The package provides a helper function for converting
back to the original scale: [convert_log2_scale_inverse](../reference/convert_log2_scale_inverse.html).**

Three further functions provide model outputs that we think are particularly useful in [data.table](https://rdatatable.gitlab.io/data.table/reference/data.table.html) format.
[biokinetics](../reference/biokinetics.html) contains documentation on each of these functions so please read that first; this vignette provides
guidance on the correct interpretation of each column in the returned tables (in these functions data is returned on the original scale).

## simulate_population_trajectories

See the documentation for this function [here](../reference/biokinetics.html#method-simulate-population-trajectories).
There are 2 different output formats depending on whether the provided `summarise` argument is `TRUE` or `FALSE`.

### summarise = TRUE

Returned columns are

| name                | type       | description                                                                                   |
|---------------------|------------|-----------------------------------------------------------------------------------------------|
| time_since_last_exp | integer    | Number of days since last exposure                                                            |
| me                  | numeric    | Median titre value                                                                            |
| lo                  | numeric    | Titre value at the 0.025 quantile                                                             |
| hi                  | numeric    | Titre value at the 0.975 quantile                                                             |
| titre_type          | character  | Name of the titre or biomarker                                                                |

There will also be a column for each covariate in the regression model.

### summarise = FALSE

Returned columns are

| name                | type    | description                        |
|---------------------|---------|------------------------------------|
| time_since_last_exp | integer | Number of days since last exposure |
| t0_pop              | numeric | Titre value at time 0              |
| tp_pop              | numeric | Time at peak titre                 |
| ts_pop              | numeric | Time at start of waning            |
| m1_pop              | numeric | Boosting rate                      |
| m2_pop              | numeric | Plateau rate                       |
| m3_pop              | numeric | Waning rate                        |
| beta_t0             | numeric | Coefficient to adjust t0 by        |
| beta_tp             | numeric | Coefficient to adjust tp by        |
| beta_ts             | numeric | Coefficient to adjust ts by        |
| beta_m1             | numeric | Coefficient to adjust m1 by        |
| beta_m2             | numeric | Coefficient to adjust m2 by        |
| beta_m3             | numeric | Coefficient to adjust m3 by        |
| mu                  | numeric | Titre value                        |
| .draw               | integer | Draw number                        |
| titre_type          | numeric | Name of the titre or biomarker     |

There will also be column for each covariate in the hierarchical model.

See the [model vignette](../article/model.html) for more detail about the model
parameters.

## simulate_individual_trajectories

See the documentation for this function [here](../reference/biokinetics.html#method-simulate-individual-trajectories).
There are 2 different output formats depending on whether the provided `summarise` argument is `TRUE` or `FALSE`.

### summarise = FALSE

Returned columns are

| name                | type                 | description                                                                                 |
|---------------------|----------------------|---------------------------------------------------------------------------------------------|
| pid                 | character or numeric | Unique person identifier as provided in input data                                          |
| draw                | integer              | Which draw from the fits this is                                                            |
| time_since_last_exp | integer              | Number of days since last exposure                                                          |
| mu                  | numeric              | Titre value                                                                                 |
| titre_type          | character            | Name of the titre or biomarker                                                              |
| exposure_day        | integer              | Day of this person's last exposure                                                          |
| calendar_day        | integer              | Day of this titre value                                                                     |
| time_shift          | integer              | The number of days these exposures have been adjusted by, as provided in function arguments |

There will also be a column for each covariate in the regression model.

### summarise = TRUE

Returned columns are

| name         | type      | description                                                                          |
|--------------|-----------|--------------------------------------------------------------------------------------|
| me           | numeric   | Median titre value                                                                   |
| lo           | numeric   | Titre value at the 0.025 quantile                                                    |
| hi           | numeric   | Titre value at the 0.075 quantile                                                    |
| titre_type   | character | Name of the titre or biomarker                                                       |
| calendar_day | integer   | Day of this titre value                                                              |
| time_shift   | integer   | The number of days the exposures were adjusted by, as provided in function arguments |

