---
title: "Data"
description: >
  Structure, prepare, transform, and inspect longitudinal biomarker data.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

`epikinetics` accepts an ordinary data frame in long format: one row per
participant, measurement date or time, and biomarker. Studies commonly record
calendar dates first and align them to an exposure-centred kinetic timescale
during preparation.

```{r raw-data, echo=FALSE, out.width="100%", fig.cap="A deterministic, history-stratified subset of nine participants from the bundled Delta-wave data. Selection spans quantiles of visit count and follow-up duration; it does not use fitted values. Colour identifies biomarker, dates show sample collection, and each dashed line marks that participant's focal exposure."}
doc_figure("documentation-longitudinal-data.png")
```

Some people contribute one visit and others several; visits occur at different
calendar dates, and each visit may include multiple biomarker targets. That
imbalance is part of the data, not something the user must regularise onto a
common grid.

## Align calendar dates

Kinetic parameters are most naturally expressed relative to an exposure, even
when the raw study records calendar dates. `align_time_to_reference()` makes
that transformation explicit and inspectable while preserving the original
dates.

```{r dated-data, collapse=FALSE}
library(epikinetics)

dated_data <- doc_delta_data()
names(dated_data)[names(dated_data) == "day"] <- "measurement_date"
names(dated_data)[names(dated_data) == "last_exp_day"] <- "exposure_date"
dated_data$measurement_date <- as.Date(dated_data$measurement_date)
dated_data$exposure_date <- as.Date(dated_data$exposure_date)

dated_data[1:5, c(
  "pid", "exposure_date", "measurement_date", "titre_type", "value"
)]
```

Here the reference is a participant-specific column. Passing `id` also checks
that its repeated value is consistent within each participant.

```{r aligned-data, collapse=FALSE}
aligned_data <- align_time_to_reference(
  dated_data,
  measurement_date = "measurement_date",
  reference = "exposure_date",
  id = "pid",
  time = "time_since_exposure"
)

aligned_data[1:5, c(
  "pid", "exposure_date", "measurement_date", "time_since_exposure"
)]
```

Use one shared date instead with, for example,
`reference = as.Date("2021-01-01")`. Differences are returned in days and may
be negative, so pre-exposure records remain visible during checking. The
current single-exposure model itself uses observations on or after time zero.

The same four participants look different before and after alignment:

```{r alignment-comparison, echo=FALSE, fig.show="hold", fig.width=4.2, fig.height=4.2, out.width="48%", fig.cap="The same observations before and after alignment. Left: participant-specific exposure dates occur at different calendar times. Right: exposure is day zero for every participant."}
alignment_participants <- doc_representative_participants(doc_delta_data(), 4L)
doc_raw_data_plot(
  doc_delta_data(), alignment_participants, aligned = FALSE, ncol = 2
) + ggplot2::theme(legend.position = "none")
doc_raw_data_plot(
  doc_delta_data(), alignment_participants, aligned = TRUE, ncol = 2
) + ggplot2::theme(legend.position = "none")
```

## Required columns

| Default column | Meaning |
|---|---|
| `pid` | User-facing participant identifier. |
| `day` | Observation time. |
| `last_exp_day` | Time of the known focal exposure. |
| `titre_type` | Biomarker or assay target. |
| `value` | Finite, positive response on the supplied scale. |

Alternative names are supplied through the `id`, `time`, `exposure`,
`biomarker`, and `value` arguments. Times may be numeric, `Date`, `POSIXt`, or
ISO date strings. Set `exposure = NULL` when `time` is already numeric time
since exposure, as in the aligned data above. The focal exposure and all model
covariates must be constant within participant.

## Prepare and inspect

```{r prepare-data, collapse=FALSE}
prepared <- prepare_epikinetics_data(
  aligned_data,
  formula = ~ infection_history,
  time = "time_since_exposure",
  exposure = NULL,
  biomarker_order = c("Ancestral", "Alpha", "Delta"),
  lower_limit = 5,
  upper_limit = 2560
)

prepared
summary(prepared)
```

The principal model components can then be inspected independently:

```{r inspect-data, collapse=FALSE}
epikinetics_data(prepared)[1:5, ]
prepared$participants[1:5, ]
model.matrix(prepared)[1:5, , drop = FALSE]
names(stan_data(prepared))
```

The prepared object retains the validated input, labelled model-ready
observations, participant table, factor levels and contrasts, biomarker
mapping, prediction grid, priors, and literal Stan data list. Integer indices
are created internally but remain visible for debugging.

## Response scale and biomarker order

With the default `scale = "natural"`, values and censoring limits are
transformed as

\[
y = \log_2(\text{value}/\text{reference value}).
\]

The fixed `reference_value` defaults to 1 and is stored for inverse
transformation. Use `scale = "log2"` only when the supplied values and limits
are already on that scale.

`biomarker_order` supplies a complete display and indexing order. Without it,
existing factor levels are preserved, then first appearance is used. The order
propagates through the fit, posterior tables, predictions, legends, and
individual plots.

Read [Covariates](covariates.html) for formula encoding and
[Censoring](censoring.html) for assay limits. Once the prepared object is
correct, continue to [Fitting the model](fitting.html).
