---
title: "Data, covariates, scales, and censoring"
description: >
  The input-data contract, transformations, formula encoding, biomarker
  ordering, censoring rules, and inspectable Stan data.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Data, covariates, scales, and censoring}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

## Required observation data

`prepare_epikinetics_data()` accepts ordinary data frames, including
`data.table` and tibble subclasses. `fit_epikinetics()` then accepts the
prepared object, keeping validation and transformation separate from
compilation and sampling. Defaults expect:

| Column | Meaning |
|---|---|
| `pid` | Participant identifier; numeric or character values are accepted. |
| `day` | Observation time. |
| `last_exp_day` | Time of the focal exposure. |
| `titre_type` | Biomarker or assay label. |
| `value` | Finite positive biomarker value on the natural scale. |

Use `id`, `time`, `exposure`, `biomarker`, and `value` arguments for other
names. Observation and exposure times may both be numeric, `Date`, `POSIXt`, or
parseable ISO date strings. The difference defines time since exposure. Times
before exposure are rejected.

```{r}
library(epikinetics)
dat <- read.csv(
  system.file("extdata", "delta.csv", package = "epikinetics")
)
dat[1:6, c("pid", "day", "last_exp_day", "titre_type", "value")]
```

The model currently represents one known focal exposure per participant, so
`last_exp_day` must be constant within participant. Missing required values and
missing covariates are rejected early.

## Scale

The Gaussian observation model operates on a base-2 logarithmic scale. With
the default `scale = "natural"`, data are transformed as

\[
y = \log_2(\text{value}/\text{reference value}).
\]

The default reference value is 1, making the inverse transformation simply
`2^y`. A scientifically meaningful assay reference can instead be supplied
with `reference_value`. This fixed reference is stored with the fit and used by
posterior summaries and response-scale predictions. It does not depend on the
minimum observed value.

For values already on this scale, use `scale = "log2"`; censoring limits must
then also be on the log2 scale. The prepared observation table retains both
`value` and `value_model`.

```{r}
prepared <- prepare_epikinetics_data(
  dat,
  formula = ~ infection_history,
  covariate_parameters = "all",
  lower_limit = 5,
  upper_limit = 2560
)
epikinetics_data(prepared)[1:6, ]
```

## Covariates

Covariates are specified with a one-sided R formula such as
`~ age + sex + infection_history`. They must be participant-level: a covariate
may not change across a participant's observations. Numeric variables remain
numeric; factors use R's active contrast options and their first level is the
default reference under treatment contrasts. Character variables are converted
to factors.

Continuous variables enter the design matrix on the scale supplied, so their
coefficients are effects per one input unit and the default coefficient priors
must be interpreted on that scale. For variables such as age, create an
explicit centred or rescaled participant-level column before preparation when
a one-unit effect is not scientifically useful. This keeps the transformation
visible in the prepared data and makes `newdata` unambiguous.

Use conventional formulas with an intercept. `epikinetics` removes the design
intercept because every biomarker already has a population parameter. Formulas
such as `~ 0 + infection_history` are rejected: estimating all factor levels in
addition to the biomarker population parameters would be redundant.

The fitted coefficients are shared across biomarkers, as in the original
model. A baseline coefficient is additive on the log2 outcome scale, so
`2^beta` is a response ratio. Coefficients for positive time and rate
parameters are additive on the parameter's log scale, so `exp(beta)` is a
ratio. `posterior_parameters(level = "regression")` returns both forms along
with the original term, numeric design-column name, represented factor level,
and treatment reference where those labels are well defined.

The original participant model frame, terms, term assignment, factor levels,
actual contrast matrices, and a design-column-to-term table are retained.
Consequently prediction never requires a user to recreate Stan dummy values.
For example, `prediction_grid(prepared)` returns the observed factor profiles
with continuous predictors fixed at their participant-level medians.

## Participant hierarchy

`participant_parameters` controls which kinetic quantities receive
participant-level random effects. The default is
`c("baseline", "boost_rate", "early_waning_rate", "late_waning_rate")`.
Peak time and the duration to the early/late waning switch are shared across
participants. This stable structure is the package default for the motivating
application, but remains an explicit and overridable modelling assumption. Use
a scientifically supported subset for another study, or
`participant_parameters = "all"` when richer within-participant sampling and
diagnostics justify the full hierarchy. The selection is inspectable in
`prepared$mappings$participant_parameters` and as the six-entry
`participant_effect_active` vector in `stan_data(prepared)`.

## Biomarker order

`biomarker_order` supplies one complete user-facing ordering and stores it in
the prepared object. If it is omitted, existing factor levels are retained;
otherwise first appearance in the input data is used. This same order controls
Stan biomarker indices, posterior tables, predictions, legends, population
plots, and individual plots. The package does not attach pathogen-specific
meaning to labels. For the bundled example, chronological variant order can be
made explicit:

```{r, eval=FALSE}
prepare_epikinetics_data(
  dat,
  biomarker_order = c("Ancestral", "Alpha", "Delta")
)
```

## Censoring

The likelihood distinguishes:

* uncensored values, using a Normal log density;
* left-censored values, using the Normal log CDF at their lower limit; and
* right-censored values, using the Normal upper-tail probability at their
  upper limit.

`lower_limit` and `upper_limit` independently accept:

* one scalar for all observations;
* a numeric vector with one value per observation;
* a named numeric vector with one limit per biomarker; or
* the name of a row-level numeric data column.

`NA` means no applicable limit for that observation. Without an explicit
`censoring` argument, a value at or below its lower limit is left-censored and
a value at or above its upper limit is right-censored. Without limits, all
values are uncensored. This means ordinary data are never implicitly censored
at their observed minimum or maximum.

If the source data record censoring status, pass a vector or column name with
labels `none`, `left`, or `right` (numeric `0`, `-1`, `1` is also accepted).
Every explicitly censored row must have the relevant limit. An observation
inconsistent with its status and applicable limit is rejected during
preparation. For example, an uncensored observation at or beyond an upper
limit, or a right-censored recorded value below that limit, produces a
row-specific error before Stan is called.

Limits can vary by biomarker:

```{r}
biomarkers <- unique(dat$titre_type)
lower <- setNames(rep(5, length(biomarkers)), biomarkers)
upper <- setNames(rep(2560, length(biomarkers)), biomarkers)

prepared <- prepare_epikinetics_data(
  dat,
  lower_limit = lower,
  upper_limit = upper
)
table(epikinetics_data(prepared)$censoring)
```

## Inspect the prepared model data

An `epikinetics_data` object keeps the validated input copy, a canonical
labelled observation table, a participant table, the R model frame, the named
participant design matrix, priors, term/design-column mappings, factor
levels/contrasts, conditional-prediction defaults, scale information, and the
exact Stan data list. Input observations are sorted contiguously by
participant for threaded likelihood evaluation without modifying the user's
data frame.

```{r}
prepared
summary(prepared)

# Validated input and model-ready observations.
prepared$input_data[1:3, ]
prepared$observations[1:3, ]

# Participant indices/covariates and biomarker indices.
prepared$participants[1:3, ]
prepared$model_frame[1:3, , drop = FALSE]
prepared$mappings$biomarker
prepared$mappings$reference_levels
prepared$mappings$design_columns

# The formula encoding and the exact list supplied to Stan.
model.matrix(prepared)[1:3, , drop = FALSE]
prediction_grid(prepared)
names(stan_data(prepared))
```

`epikinetics_data(prepared)` is a compact accessor for the observation table,
whereas `stan_data(prepared)` returns the exact sampling list. The direct
`prepared$stan_data` component is also
available for interactive debugging. Integer indices need not be supplied by
the user, but they are intentionally visible. The fit updates only the
computational `grainsize` for the requested thread count; after sampling,
`stan_data(fit)` is the literal list used for that run.

## Fit and posterior output

`fit_epikinetics()` returns one `epikinetics_fit` object containing the
CmdStanR fit and all preparation metadata needed for later interpretation. The
underlying `CmdStanMCMC` object remains available through `cmdstan_fit(fit)`;
raw Stan draws are available through `posterior_draws(fit)`.

`posterior_parameters()` provides labelled kinetic quantities at four levels:

| `level` | Interpretation | Main identifiers |
|---|---|---|
| `"population"` | Biomarker-specific population intercept parameters before covariate effects. | `biomarker` |
| `"profile"` | Population parameters after applying a conditional covariate profile, without participant effects. | `.profile`, `biomarker`, and formula variables |
| `"participant"` | Conditional parameters for fitted participants, including their estimated random effects. | `participant`, `biomarker`, and participant covariates |
| `"regression"` | Formula coefficients and their interpretable multiplicative forms. | formula term, design column, level/reference, and kinetic parameter |

With `summary = TRUE`, parameter output is long: `parameter` (or `quantity`
for regression output) names the quantity and `mean`, `median`, `sd`, `lower`,
and `upper` give its posterior summary. With `summary = FALSE`, one row is
returned per posterior draw and the kinetic quantities occupy separate
columns. Derived output includes:

| Column | Meaning |
|---|---|
| `waning_change_time` | Time of the transition from early to late waning. |
| `baseline_response` | Baseline on the natural response scale. |
| `peak_response` | Value at the fitted peak on the response scale. |
| `waning_change_response` | Value at the early/late waning switch. |
| `early_waning_half_life` | Time for a two-fold decline under the early rate. |
| `late_waning_half_life` | Time for a two-fold decline under the late rate. |

The old package referred to `waning_change_response` as a set point. It is not
necessarily stationary: the late waning rate is estimated and may be positive.

## Trajectory-prediction output

`predict(fit)` returns an `epikinetics_prediction`, which is an ordinary data
frame with plotting metadata attached. Every summarised prediction contains:

| Column | Meaning |
|---|---|
| `time` | Time since the focal exposure, in the input time unit. |
| `biomarker` | User-facing biomarker label in the stored order. |
| `estimate` | Alias for the posterior median. |
| `mean` | Posterior mean trajectory value. |
| `median` | Posterior median trajectory value. |
| `lower`, `upper` | Requested pointwise posterior interval. |

Population and new-participant predictions also retain `.profile` and the
original covariate-profile columns. Individual predictions retain
`participant` and the fitted participant's covariate columns. All outputs are
on the response scale by default; `scale = "model"` requests the fitted log2
scale.

With `summary = FALSE`, the output instead contains `.draw`, `time`, and
`estimate` together with the same type-specific identifiers. Keeping `.draw`
is essential when calculating quantities involving more than one time point,
biomarker, or participant: summaries must be formed across complete
draw-specific calculations rather than by combining marginal medians.

The prediction `type` determines the estimand:

| `type` | Random effects included |
|---|---|
| `"population"` | None; conditional on the supplied or default covariate profiles. |
| `"individual"` | Posterior effects for fitted participants. |
| `"new"` | Newly drawn effects from the fitted participant hierarchy. |

Intervals describe uncertainty in the latent expected trajectory unless
`include_observation_noise = TRUE`, in which case they are posterior
predictive intervals for a future observed measurement. Attributes record the
interval probabilities, scale, uncertainty target, censoring limits,
biomarker order, formula, prediction grid, and stored observations used by the
plot methods.

The applied case-study vignette shows how draw-level individual predictions
can be placed on calendar time and aggregated without losing draw identity.

The package bundles `delta.csv`, `ba2.csv`, and `xbb.csv` under `extdata` for
examples and tests. They are observational examples from the motivating study,
not implicit defaults for modelling or priors.
