---
title: "Model diagnostics and plotting"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{Model diagnostics and plotting}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Plotting model inputs

### Priors
Calling `plot` on an object of type [biokinetics_priors](../reference/biokinetics_priors.html) will generate
a plot of the kinetics predicted by the given priors.

```{r}
priors <- epikinetics::biokinetics_priors()
plot(priors)
```

You can optionally pass a dataset to compare against the predicted kinetics. Required columns are
`time_since_last_exp` and `value` and values should be on the scale required by the model.

```{r}
data <- data.table::fread(system.file("delta_full.rds", package = "epikinetics"))
data[, time_since_last_exp := as.integer(day - last_exp_day, units = "days")]
data <- epikinetics::convert_log2_scale(data, min(data$value))
priors <- epikinetics::biokinetics_priors()
plot(priors, data = data)
```

If you have an instance of the [biokinetics](../reference/biokinetics.html) class, the method `plot_prior_predictive`
generates this plot for the priors and data given to the model.

```{r}
data <- data.table::fread(system.file("delta_full.rds", package = "epikinetics"))
priors <- epikinetics::biokinetics_priors()
mod <- epikinetics::biokinetics$new(priors = priors, data = data)
mod$plot_prior_predictive()
```

### Data

If you have an instance of the [biokinetics](../reference/biokinetics.html) class, the method `plot_model_inputs`
plots the input data used to fit the model, disaggregated by the covariates in the covariate formula.

```{r}
mod <- epikinetics::biokinetics$new(priors = priors, data = data, covariate_formula = ~0 + infection_history)
mod$plot_model_inputs()
```

## Interactive data exploration

To play around with different priors and visualise input data filtered and disaggregated in different ways,
the function [biokinetics$inspect](reference/biokinetics.html#method-biokinetics-inspect) runs a local RShiny app with interactive plots.

![RShiny demonstration](./shiny.webm)
