---
title: "Diagnostics and posterior prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics and posterior prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Before fitting

Input and prior plots are intentionally independent of the sampler. They return
ordinary `ggplot` objects and can be customised with ggplot2.

```{r}
library(epikinetics)
dat <- read.csv(
  system.file("extdata", "delta.csv", package = "epikinetics")
)
prepared <- prepare_epikinetics_data(
  dat,
  formula = ~ infection_history,
  lower_limit = 5,
  upper_limit = 2560
)

summary(prepared)
model.matrix(prepared)[1:5, , drop = FALSE]
prepared$mappings$participant[1:5, ]
table(prepared$observations$censoring)
names(stan_data(prepared))

plot(prepared)
plot(epikinetics_priors(), scale = "response", reference_value = 1)
```

Prior trajectory plots are a quick scale check, not a substitute for a full
prior predictive workflow. In particular, they draw population trajectories
without participant or observation variation.

## Sampling diagnostics

The following examples assume a fit created as in the getting-started
vignette. They are not evaluated when the vignette is built.

```{r, eval=FALSE}
diagnostics <- diagnose_epikinetics(fit)
diagnostics$overview
diagnostics$sampler
diagnostics$chains
diagnostics$parameters
```

The overview reports:

* divergent transitions;
* transitions hitting maximum tree depth;
* chains with non-finite E-BFMI, separately from chains below 0.3;
* maximum R-hat among principal population parameters; and
* minimum bulk and tail effective sample sizes for those parameters.

These are screening summaries. Inspect parameter-level diagnostics and trace
plots whenever the screening result is poor. The full CmdStanR object is
available without copying draws:

```{r, eval=FALSE}
stan_fit <- cmdstan_fit(fit)
stan_fit$diagnostic_summary()
stan_fit$cmdstan_diagnose()
posterior::as_draws_df(stan_fit$draws())
```

`diagnostics$chains` also reports the number and variance of retained energy
values. For example, `E-BFMI = NaN` together with one unique energy value means
the chain was frozen: both the numerator and denominator of the E-BFMI ratio
are zero. It is not a benign missing diagnostic. Predictions keep all retained
chains, so they warn when E-BFMI is non-finite instead of silently selecting the
apparently better chains.

If all or some chains fail, `fit_epikinetics()` preserves the CmdStanR object
and marks the returned object as failed or partial whenever CmdStanR supplies a
run object. The original chain output remains the primary debugging record:

```{r, eval=FALSE}
fit$computation$sampling_state
cmdstan_fit(fit)$return_codes()
cmdstan_fit(fit)$output()
```

This complements inspection of `stan_data(prepared)`; it does not wrap or
replace CmdStanR's diagnostic interface.

The previous package's Shiny inspector explored inputs and priors but was not a
posterior diagnostic application. It has been removed from the core interface:
the lighter plotting functions cover its maintainable role, and CmdStanR plus
the posterior ecosystem provide more complete sampler diagnostics.

## Latent trajectories versus observations

By default, `predict()` returns the latent expected biomarker curve, excluding
residual measurement noise. Set `include_observation_noise = TRUE` when the
target is a future observed measurement rather than its latent mean.

```{r, eval=FALSE}
latent <- predict(
  fit,
  times = 0:150,
  type = "population",
  summary = FALSE,
  ndraws = 500
)

observed <- predict(
  fit,
  times = 0:150,
  type = "population",
  include_observation_noise = TRUE,
  summary = FALSE,
  ndraws = 500,
  seed = 10
)
```

Keeping `summary = FALSE` preserves the `.draw` identifier, which should be
retained when combining trajectory quantities across times or biomarkers.
With `summary = TRUE`, `lower` and `upper` are calculated across these raw
draw-specific trajectories at each biomarker/time/profile combination. The
plot ribbon uses those columns directly.

## Prediction targets

Population predictions omit participant random effects. When `newdata` is
`NULL`, `prediction_grid()` supplies the combinations of categorical values
observed among fitted participants and fixes continuous predictors at their
participant-level medians. This avoids silent extrapolation to impossible
factor combinations. It is a conditional estimand, not marginalisation over
the sample's covariate distribution.

```{r, eval=FALSE}
prediction_grid(fit)
population <- predict(fit, type = "population")
plot(population)

# Explicit newdata selects or overrides the conditional profiles.
profiles <- data.frame(
  infection_history = c(
    "Infection naive",
    "Previously infected (Pre-Omicron)"
  )
)
population <- predict(fit, newdata = profiles, type = "population")
plot(population)
```

For multiple categorical predictors the default includes only observed
combinations. `prediction_grid(fit, categorical = "cartesian")` deliberately
requests every fitted-level combination. The stored R terms object and
contrasts evaluate main effects, interactions, and transformed terms for both
automatic and explicit grids. Unknown factor levels fail before posterior
prediction with an error naming the variable, supplied level, and fitted
levels.

Existing-participant predictions combine population, covariate, and fitted
random effects:

```{r, eval=FALSE}
individual <- predict(
  fit,
  type = "individual",
  participants = c("1", "2"),
  times = 0:180
)
plot(individual)
```

New-participant predictions draw one new standardised random effect per
*selected* participant-level kinetic quantity and posterior draw. Inactive
quantities remain at their conditional population/profile values. Each active
effect is used across biomarkers, in accordance with the fitted hierarchy:

```{r, eval=FALSE}
new_individuals <- predict(
  fit,
  newdata = profiles,
  type = "new",
  times = 0:180,
  seed = 34
)
plot(new_individuals)
```

These new-participant results represent one hypothetical participant per
profile per draw. They are not an average over a newly simulated cohort.

## Scales and plots

Response-scale output is the default and applies the reference stored during
data preparation. Use `scale = "model"` for log2 output. `plot(fit)` produces
formula-aware conditional population trajectories with observations overlaid,
while
`plot(fit, type = "data")` shows prepared inputs only.

```{r, eval=FALSE}
plot(fit)
plot(fit, type = "data")

p <- plot(predict(fit, type = "population", times = 0:180))
p + ggplot2::theme_minimal()
```

Plot methods do no scientific aggregation beyond the documented posterior
summary. For custom analyses, work with the returned prediction data frame or
raw draws directly.
