---
title: "Diagnostics"
description: >
  Screen sampling behaviour and inspect the underlying CmdStanR fit before
  interpreting posterior results.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Diagnostics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

Diagnostics answer whether the sampler explored a common, stable posterior.
Check them before interpreting parameters or trajectories.

```{r, eval=FALSE}
diagnostics <- diagnose_epikinetics(fit)
diagnostics$overview
diagnostics$chains
diagnostics$parameters
```

```{r, echo=FALSE, out.width="96%", fig.cap="Representative trace and marginal-density panels from the four-chain documentation fit. Each column isolates one chain; corresponding density shapes provide a compact check that chains explore the same stationary distribution. Numerical diagnostics remain essential alongside this visual screen."}
doc_figure("documentation-diagnostics.png")
```

The two parameters deliberately represent different kinetic features: the
Ancestral population time to peak and its late waning rate. Showing every chain
separately avoids hiding a poorly mixing chain behind an overlaid trace, while
the matched density row makes between-chain agreement easy to compare.

## What to check

The compact report screens:

* **divergent transitions**, which indicate that HMC could not reliably explore
  part of the posterior geometry;
* **maximum-treedepth hits**, which indicate trajectories were truncated;
* **R-hat**, which should be close to 1 for each estimand;
* **bulk and tail effective sample sizes**, which quantify information in the
  autocorrelated draws; and
* **E-BFMI by chain**, which checks whether HMC explored the energy distribution.

A non-finite E-BFMI is not treated as a harmless missing value. The chain table
also reports retained energy counts and variance, helping distinguish a frozen
chain with constant energy from ordinary low E-BFMI.

The report is a screen, not a replacement for parameter-level inspection.
Use the underlying CmdStanR and posterior interfaces for deeper work:

```{r, eval=FALSE}
stan_fit <- cmdstan_fit(fit)
stan_fit$diagnostic_summary()
stan_fit$cmdstan_diagnose()
draws <- posterior::as_draws_df(stan_fit$draws())
```

If a chain fails, inspect its original process output:

```{r, eval=FALSE}
fit$computation$sampling_state
cmdstan_fit(fit)$return_codes()
cmdstan_fit(fit)$output()
```

Do not discard a problematic chain merely to improve a summary. Diagnose data,
initialisation, model geometry, and the affected parameters. The
[Stan diagnostics guide](https://mc-stan.org/learn-stan/diagnostics-warnings.html)
provides the broader interpretation of each warning.

Once sampling is trustworthy, continue to
[Population-level kinetics](population-kinetics.html),
[Individual-level kinetics](individual-kinetics.html), or the
[case study](case-study.html).
