---
title: "Replicating Figure 2"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{delta}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Figure 2 from the paper shows population level fits for each wave, disaggregated by infection history
and titre type. This is a partial reconstruction of that figure to demonstrate how to use the package.

To generate this figure, first run the model for each wave separately, using the data files installed with this package.
Here we run it just for the Delta wave:

```{r}
dat <- data.table::fread(system.file("delta_full.rds", package = "epikinetics"))
mod <- epikinetics::sam$new(data = dat)
delta <- mod$fit()
```

Once the model has been fitted, process the fits into trajectories:

```{r}
res <- mod$population_trajectories()
head(res)
```

And now plot using `ggplot`:

```{r}
library(ggplot2)
manual_pal <-
  c("#CC6677",
    "#DDCC77",
    "#88CCEE",
    "#882255",
    "#44AA99",
    "grey",
    "#D95F02",
    "#66A61E")
res$titre_type <- forcats::fct_relevel(res$titre_type, c("Ancestral", "Alpha", "Delta"))
ggplot() +
  geom_line(data = res,
            aes(x = t,
                y = me,
                colour = titre_type,
                linetype = "Retrospective")) +
  geom_ribbon(data = res,
              aes(x = t,
                  ymin = lo,
                  ymax = hi,
                  fill = titre_type), alpha = 0.65) +
  scale_y_continuous(
    trans = "log2",
    breaks = c(40, 80, 160, 320, 640, 1280,
               2560, 5120),
    labels = c("40", "80", "160", "320", "640", "1280", "2560", "5120"),
    limits = c(40, 10240)) +
  scale_x_continuous(breaks = c(0, 30, 60, 90, 120),
                     labels = c("0", "30", "60", "90", "120"),
                     expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  labs(x = "Time since last exposure (days)",
       y = expression(paste("Titre (IC"[50], ")"))) +
  facet_wrap(infection_history ~ titre_type) +
  theme(
    legend.position = "bottom",
    strip.text.x.top = element_text(size = 8, family = "Helvetica"),
    strip.text.x = element_text(size = 8, family = "Helvetica")) +
  scale_colour_manual(values = manual_pal) +
  scale_fill_manual(values = manual_pal) +
  theme_linedraw() +
  theme(legend.position = "bottom",
        text = element_text(size = 8, family = "Helvetica"),
        strip.background = element_rect(fill = "white"),
        strip.text = element_text(colour = 'black'),
        strip.placement = "outside",
        plot.title = element_text(face = "bold", size = 9),
        panel.grid = element_line(linewidth = 0.4)) +
  guides(colour = "none", fill = "none")
```
