---
title: "Fitting the model"
description: >
  Compile and sample the Stan model with parallel chains and within-chain
  threading.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Fitting the model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

`fit_epikinetics()` receives a prepared-data object, configures CmdStanR, and
returns one fit containing the posterior run plus everything needed for
interpretation and prediction.

```{r, echo=FALSE, out.width="96%", fig.cap="Parallel chains are independent MCMC runs. Within each chain, Stan's reduce_sum likelihood partitions observations by participant across worker threads."}
doc_figure("documentation-sampling.svg")
```

## Run the sampler

```{r, eval=FALSE}
fit <- fit_epikinetics(
  prepared,
  chains = 4,
  parallel_chains = 4,
  threads_per_chain = 2,
  iter_warmup = 1000,
  iter_sampling = 1000,
  adapt_delta = 0.9,
  seed = 2026
)
```

`parallel_chains` controls how many chains run simultaneously.
`threads_per_chain` controls workers within each chain; the model is always
compiled with `STAN_THREADS`, and its likelihood is partitioned by participant
through `reduce_sum`. The automatic grainsize is usually a good starting
point; change it only after benchmarking a representative dataset.

`adapt_delta`, `max_treedepth`, and additional CmdStanR sampling arguments
remain available. More conservative controls should respond to diagnosed
geometry rather than replace model checking.

## Compilation and CmdStan access

Package installation and loading do not install CmdStan or compile a model.
Install CmdStan explicitly with:

```{r, eval=FALSE}
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
```

The first fit compiles a threaded executable in the user's R cache; matching
later fits reuse it. Missing CmdStan produces a setup-oriented error rather
than an attempted automatic installation.

Advanced users can use the original CmdStanR object directly:

```{r, eval=FALSE}
stan_fit <- cmdstan_fit(fit)
stan_fit$summary()
stan_fit$diagnostic_summary()
posterior_draws(fit)
```

If chains fail, the package preserves the CmdStanR run whenever one exists:

```{r, eval=FALSE}
cmdstan_fit(fit)$return_codes()
cmdstan_fit(fit)$output()
```

Always continue to [Diagnostics](diagnostics.html) before interpreting
parameters or trajectories.
