---
title: "Getting started with epikinetics"
description: >
  Prepare longitudinal biomarker data, fit the model, check diagnostics, and
  plot conditional population kinetics.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Getting started with epikinetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

`epikinetics` fits hierarchical kinetics models to repeated positive
biomarker measurements after one focal exposure. This page is the shortest path
from a data frame to a fitted population trajectory; the focused articles
linked at the end explain each step.

## Prepare the data

The package bundles public neutralising-antibody data for examples. Preparation
is deliberately separate from fitting, so every transformation can be checked
before starting Stan.

```{r, collapse=FALSE}
library(epikinetics)

dat <- read.csv(
  system.file("extdata", "delta.csv", package = "epikinetics")
)

prepared <- prepare_epikinetics_data(
  dat,
  formula = ~ infection_history,
  biomarker_order = c("Ancestral", "Alpha", "Delta"),
  lower_limit = 5,
  upper_limit = 2560
)

prepared
prediction_grid(prepared)
```

Use `summary(prepared)`, `model.matrix(prepared)`, and
`stan_data(prepared)` to inspect the validated observations, formula encoding,
participant/biomarker indices, censoring limits, and exact Stan data. The
[Data](data.html), [Covariates](covariates.html), and
[Censoring](censoring.html) articles describe those contracts.

## Fit and diagnose

CmdStan is installed explicitly once with `cmdstanr::install_cmdstan()`. The
first fit compiles and caches the threaded model; later fits reuse it.

```{r, eval=FALSE}
fit <- fit_epikinetics(
  prepared,
  chains = 4,
  parallel_chains = 4,
  threads_per_chain = 2,
  seed = 2026
)

fit
diagnose_epikinetics(fit)
```

Do not interpret posterior estimates until the diagnostic summary is
satisfactory. [Fitting the model](fitting.html) explains the computational
arguments, and [Diagnostics](diagnostics.html) explains the checks.

## Predict population kinetics

```{r, eval=FALSE}
pred <- predict(fit, type = "population", times = 0:150)
plot(pred)
```

```{r, echo=FALSE, out.width="96%", fig.cap="Conditional population kinetics produced by the code above. Biomarkers are overlaid by colour; infection-history profiles define panels. Lines are posterior medians and ribbons are 95% credible intervals for the latent trajectories."}
doc_figure("documentation-population-kinetics.png")
```

The figure is precomputed from an actual four-chain package fit so routine
documentation builds do not rerun MCMC. Both posterior mean and median remain
in `pred`; the plotting method uses the median by default.

Continue with [Population-level kinetics](population-kinetics.html) for
prediction grids, `newdata`, and uncertainty targets, or
[Individual-level kinetics](individual-kinetics.html) to reconstruct fitted
participants. The [case study](case-study.html) then combines the workflow in a
complete applied analysis.
