---
title: "Getting started with epikinetics"
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Getting started with epikinetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

`epikinetics` fits a Bayesian hierarchical kinetics model to repeated biomarker
measurements after one focal exposure. This vignette shows the intended path
from a data frame to diagnostics and posterior trajectories. The fitting chunks
are not evaluated while building the vignette because they require a local
CmdStan installation.

## Set up CmdStan

CmdStanR and CmdStan are external to the package. Install CmdStan explicitly
once:

```{r, eval=FALSE}
cmdstanr::check_cmdstan_toolchain(fix = TRUE)
cmdstanr::install_cmdstan()
cmdstanr::cmdstan_version()
```

The first model fit compiles a threaded executable and caches it. Package
installation and loading do not compile a model or install CmdStan.

## Read, prepare, and inspect data

The package includes public example data from the motivating SARS-CoV-2 study.
They are ordinary CSV files rather than special R objects.

```{r}
library(epikinetics)

dat <- read.csv(
  system.file("extdata", "delta.csv", package = "epikinetics")
)
dat[1:4, c("pid", "day", "last_exp_day", "titre_type", "value",
           "infection_history")]
```

Validation and transformation can be run separately. This is fast, does not
need CmdStan, and makes scale, censoring, indices, and the model matrix
inspectable before sampling.

```{r prepare-data, fig.cap="Prepared observations on the response scale. Biomarkers retain the order supplied during data preparation, and point shapes distinguish censoring status."}
prepared <- prepare_epikinetics_data(
  dat,
  formula = ~ infection_history,
  covariate_parameters = "all",
  biomarker_order = c("Ancestral", "Alpha", "Delta"),
  lower_limit = 5,
  upper_limit = 2560
)
prepared
summary(prepared)
plot(prepared)

# These are the transformed observations, participant mapping, formula
# encoding, and exact list that will be supplied to Stan.
prepared$observations[1:6, ]
prepared$participants[1:6, ]
prepared$mappings$biomarker
prepared$mappings$reference_levels
prepared$mappings$design_columns
model.matrix(prepared)[1:6, , drop = FALSE]
str(stan_data(prepared), max.level = 1)
```

The formula uses normal R treatment contrasts. Its intercept is required in
the formula and removed internally because biomarker-specific population
parameters already act as intercepts. Here, `infection_history` must be
constant within each participant.

This separation is deliberate. Longitudinal ordering, response-scale
transformation, censoring, participant indices, and factor contrasts can all
change the statistical input. They should be inspectable before running a
computationally expensive Bayesian model. `fit_epikinetics()` therefore
requires a prepared object rather than silently preparing a raw data frame.

## Fit

```{r, eval=FALSE}
fit <- fit_epikinetics(
  prepared,
  chains = 4,
  parallel_chains = 4,
  threads_per_chain = 2,
  iter_warmup = 1000,
  iter_sampling = 1000,
  seed = 2026
)
fit
summary(fit)
```

Parallel chains run independent chains simultaneously. Threads within a chain
evaluate participant groups through Stan's `reduce_sum`. The default grainsize
is based on the number of participants and threads, and can be overridden with
`grainsize` after benchmarking a representative dataset.

## Diagnose

Always inspect diagnostics before interpreting the posterior.

```{r, eval=FALSE}
diagnostics <- diagnose_epikinetics(fit)
diagnostics$overview

# Optionally run CmdStan's full text diagnostic utility too.
diagnose_epikinetics(fit, run_cmdstan = TRUE)
```

The compact report covers divergences, maximum-treedepth hits, E-BFMI, R-hat,
and effective sample sizes. The underlying CmdStanR fit remains available for
all standard diagnostics:

```{r, eval=FALSE}
stan_fit <- cmdstan_fit(fit)
stan_fit$diagnostic_summary()
stan_fit$summary()
```

## Extract parameters

```{r, eval=FALSE}
posterior_parameters(fit, level = "population")
posterior_parameters(fit, level = "regression")
posterior_parameters(fit, level = "profile")
posterior_parameters(
  fit,
  level = "participant",
  participants = c("1", "2")
)
```

Population and participant output includes the six fitted kinetic parameters
and, by default, interpretable derived quantities: transition time, peak and
transition values, and early/late waning half-lives. For a log2-linear waning
rate (r), half-life is (1/r) in the time unit of the input. Set
`summary = FALSE` to retain one row per posterior draw.

`level = "profile"` applies fitted regression effects to the same automatic
conditional grid used by `predict()`, retaining biomarker and original
covariate labels. `level = "population"` instead reports the underlying
biomarker-specific intercept parameters before covariate effects.

Raw Stan draws can be obtained without translating parameter names:

```{r, eval=FALSE}
draws <- posterior_draws(fit)
```

## Predict and plot trajectories

The three prediction types answer distinct questions:

* `"population"` applies covariates but excludes participant variation;
* `"individual"` reconstructs fitted participants using their posterior
  effects; and
* `"new"` draws effects for new participants with supplied covariates.

By default, participant effects are fitted for baseline, boost rate, early
waning rate, and late waning rate. Peak timing and the duration from peak to the
early/late waning switch have no residual participant random effect after
conditioning on selected covariates. This is an explicit modelling assumption
that is appropriate and stable for the motivating data, not a universal
immunological claim. Select another scientifically supported subset with
`participant_parameters` during preparation; new-participant predictions draw
only the selected effects.

```{r, eval=FALSE}
# The formula is carried forward automatically. This grid contains the
# infection-history categories observed in the fitted participant data.
prediction_grid(fit)

population <- predict(
  fit,
  times = 0:150,
  type = "population"
)
plot(population)

# Override the grid with any supported fitted levels.
profiles <- data.frame(
  infection_history = c(
    "Infection naive",
    "Previously infected (Pre-Omicron)"
  )
)
selected_population <- predict(fit, newdata = profiles, times = 0:150)

individual <- predict(
  fit,
  participants = c("1", "2"),
  times = 0:150,
  type = "individual",
  ndraws = 500
)
plot_individual(individual, participant = "1")

# Equivalent one-participant convenience method.
plot_individual(fit, participant = "1")

new_people <- predict(
  fit,
  newdata = profiles,
  times = 0:150,
  type = "new",
  seed = 14
)
plot(new_people)

# Convenient formula-aware population plot with observed data overlaid.
plot(fit)

# Reuse one bounded-memory summary for a cohort PDF.
all_individuals <- predict(
  fit,
  type = "individual",
  times = 0:150,
  ndraws = 500
)
save_individual_plots(
  all_individuals,
  path = "individual-plots",
  format = "pdf",
  multipage = TRUE
)
```

Default profiles use observed combinations of categorical variables and hold
continuous variables at their participant-level medians. They are conditional
trajectories for those profiles, not covariate-marginal averages.
`prediction_grid()` exposes the inputs before prediction and can explicitly
construct a Cartesian factor grid when extrapolation to unobserved
combinations is scientifically justified. The original terms object,
contrasts, levels, interactions, and transformations are reused to build each
numeric design row.

Prediction summaries retain both posterior `mean` and `median`; plots use the
median by default and accept `central = "mean"`. Population figures overlay
biomarkers by colour and facet over categorical profiles derived from the
stored formula. Censoring limits are dashed lines, and response-scale figures
use log2 spacing with natural-scale labels. The order supplied through
`biomarker_order` is retained in every output and legend; without one, factor
levels and then first appearance determine the order.

Individual trajectories are evaluated in R from posterior kinetic-parameter
draws. Summary mode processes participants in chunks rather than constructing
one enormous draw-by-participant-by-biomarker-by-time data frame, while
`participants`, `biomarkers`, `times`, and `ndraws` provide explicit controls.
This also allows arbitrary time grids after fitting and avoids large generated-
quantities output. Use `summary = FALSE` only for deliberately sized draw-level
output.

The default interval describes uncertainty in the latent mean trajectory. Use
`include_observation_noise = TRUE` for an explicitly labelled posterior
predictive interval for a future measurement. All plotting methods return
ordinary `ggplot` objects.

The data, model, and diagnostics vignettes provide the full contracts and
statistical assumptions behind this workflow. The applied case-study vignette
uses the bundled Delta-wave data to reconstruct the population, peak/switch,
calendar-time cohort, and exposure-timing analyses from the original package.
