---
title: "Individual-level kinetics"
description: >
  Reconstruct fitted participants, compare latent curves with observations,
  and export participant plots.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Individual-level kinetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

Individual predictions combine population kinetics, covariate effects, and a
fitted participant's posterior random effects. Original participant identifiers
are retained throughout.

```{r, eval=FALSE}
individual <- predict(
  fit,
  type = "individual",
  participants = "202",
  times = 0:220,
  ndraws = 1000
)

plot_individual(individual, participant = "202")
```

```{r, echo=FALSE, out.width="96%", fig.cap="Participant 202, the most repeatedly observed member of the deterministic raw-data display subset. Biomarkers retain their established colours but use separate facets to prevent overlapping trajectories. Points are measurements, lines are posterior median latent trajectories, and ribbons are pointwise 95% credible intervals."}
doc_figure("documentation-individual-kinetics.png")
```

The participant was selected using only the raw sampling design, not model fit
quality. Faceting makes biomarker-specific fit easier to inspect without losing
the colour, censoring, response-scale, or ordering conventions used in the
population plot.

## Select output deliberately

```{r, eval=FALSE}
selected <- predict(
  fit,
  type = "individual",
  participants = c("31", "94", "202"),
  biomarkers = c("Ancestral", "Delta"),
  times = seq(0, 180, by = 3),
  ndraws = 500
)
plot(selected)
```

Summary mode processes participants in bounded chunks, avoiding one enormous
draw × participant × biomarker × time table. `participants`, `biomarkers`,
`times`, `ndraws`, and `chunk_size` give explicit control. Unsummarised
output is protected by `max_rows`; increase it only after considering memory.

Prediction is performed in vectorised R code from posterior kinetic-parameter
draws. This permits arbitrary post-fit time grids and avoids storing huge
generated-quantity arrays during sampling. Package tests compare the R kinetic
function numerically with the Stan implementation.

## New participants and batch plots

`type = "new"` draws new effects from the fitted hierarchy for supplied
covariate profiles. It is different from reconstructing a fitted person:

```{r, eval=FALSE}
new_people <- predict(
  fit,
  type = "new",
  newdata = data.frame(infection_history = "Infection naive"),
  times = 0:180,
  seed = 34
)
```

For cohort review, compute one reusable prediction and save PNG files,
individual PDFs, or one multi-page PDF:

```{r, eval=FALSE}
all_individuals <- predict(
  fit,
  type = "individual",
  times = 0:220,
  ndraws = 500
)

save_individual_plots(
  all_individuals,
  path = "individual-plots",
  format = "pdf",
  multipage = TRUE
)
```

The default band is uncertainty in the latent curve; request observation noise
only when the estimand is a future measurement. Read [Diagnostics](diagnostics.html)
before using individual reconstructions substantively.
