---
title: "Statistical model and priors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Statistical model and priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Scope

The model describes repeated measurements after one focal exposure. It keeps
the scientific structure of the original `epikinetics` model: a piecewise
linear trajectory on the log2 scale, biomarker-specific population kinetics,
participant variation, participant covariates, Gaussian observation error, and
left/right censoring. It does not model repeated exposures or infer exposure
times.

The implementation uses explicit names in R and Stan. The correspondence to
the original mathematical notation is:

| Public name | Original name | Interpretation |
|---|---:|---|
| `baseline` | `t0` | Expected log2 value at exposure time. |
| `time_to_peak` | `tp` | Positive time from exposure to the peak. |
| `waning_duration` | `ts - tp` | Positive duration of early waning. |
| `boost_rate` | `m1` | Positive pre-peak slope. |
| `early_waning_rate` | `-m2` | Positive magnitude of the first post-peak decline. |
| `late_waning_rate` | `-m3` | Positive magnitude after the waning-rate transition. |

`waning_change_time = time_to_peak + waning_duration` corresponds to `ts`.

## Kinetic curve

For a participant-biomarker pair, let baseline be (b), time to peak (t_p),
waning-rate change time (t_s > t_p), boost rate (r_b > 0), early waning
rate (r_e > 0), and late waning rate (r_l > 0). The latent log2 mean is

\[
\mu(t) =
\begin{cases}
b + r_b t, & t \le t_p, \\
b + r_b t_p - r_e(t-t_p), & t_p < t \le t_s, \\
b + r_b t_p - r_e(t_s-t_p) - r_l(t-t_s), & t > t_s.
\end{cases}
\]

The curve is continuous. The second post-peak rate is not structurally zero:
it describes long-term waning, and a posterior near zero represents a plateau.
This is the behaviour of the original Stan implementation and its defaults,
despite older documentation calling the interval from `tp` to `ts` a plateau
and saying `m2` was fixed at zero. The clearer interpretation is boost, early
waning, then late waning or plateau.

There is no artificial floor at zero on the model scale. The former
`max(mu, 0)` caused a non-differentiable flat region and made the implied floor
depend on a data-derived reference. With a fixed response transformation,
log2 values below zero are valid positive response values below the reference.

## Population, covariate, and participant effects

Each biomarker (k) has its own population baseline and five positive kinetic
parameters. Let (q_k) be a positive population time or rate, (x_i) the
participant design row, and (z_{q,i}\sim\mathrm{Normal}(0,1)). Participant
parameters are

\[
b_{ik} = b_k + x_i^T\beta_b + \sigma_{b,k}z_{b,i}
\]

and

\[
q_{ik} = q_k\exp(x_i^T\beta_q + \sigma_{q,k}z_{q,i}).
\]

Thus baseline effects are additive in log2 units, whereas positive time and
rate effects are multiplicative. This guarantees positive rates, positive
times, and (t_s > t_p) for every participant and covariate profile. The old
additive construction could reverse rate signs or produce an individual
change time before their peak.

Active random effects use a non-centred parameterisation. The standardised effect
(z_{q,i}) is shared across biomarkers and scaled by a biomarker-specific
standard deviation. This preserves an original scientific assumption: a
participant high on a kinetic quantity for one biomarker tends to be high on
that quantity for all biomarkers. A future model could estimate a richer
cross-biomarker correlation structure, but this pass does not add it.

The default participant hierarchy includes `baseline`, `boost_rate`,
`early_waning_rate`, and `late_waning_rate`. Participants may therefore differ
in their baseline response, induction slope, and both waning components. Peak
timing and `waning_duration` (the positive duration from peak to the switch in
waning rate) have no residual participant random effect by default after
conditioning on selected covariates. This is a modelling assumption supported
by the motivating data and its posterior geometry, not a claim that transition
timing can never vary biologically. Another dataset can
opt into a different selected hierarchy or all six effects:

```{r, eval=FALSE}
prepare_epikinetics_data(
  dat,
  participant_parameters = c(
    "baseline", "boost_rate", "early_waning_rate", "late_waning_rate"
  )
)

prepare_epikinetics_data(dat, participant_parameters = "all")
```

Inactive effects create neither latent standard-Normal parameters nor
hierarchical SD parameters in Stan. Their reported participant SD is zero, and
participant/new-participant predictions vary only over the fitted hierarchy.

Covariate coefficient vectors are also shared across biomarkers, preserving the
original hierarchy. R constructs a conventional participant-level design
matrix and records its contrasts so new profiles are encoded consistently.
The default applies that matrix to every kinetic parameter, as the legacy model
did, but the choice is no longer implicit. `covariate_parameters` can name a
scientifically motivated subset:

```{r, eval=FALSE}
prepare_epikinetics_data(
  dat,
  formula = ~ infection_history,
  covariate_parameters = c("baseline", "boost_rate", "late_waning_rate")
)
```

An active coefficient for baseline is an additive change in expected log2
level. For every positive time or rate it is a log ratio: exponentiating the
coefficient gives the multiplicative population shift associated with a
one-unit design-matrix change. Inactive parameter/formula combinations do not
create unused Stan coefficients.

## Observation and censoring model

For an uncensored measurement on the model scale,

\[
y_n \sim \mathrm{Normal}(\mu_n,\sigma_{obs}).
\]

The observation SD is common to biomarkers, as in the original model. A
left-censored observation contributes
`normal_lcdf(lower_limit | mu, observation_sd)` and a right-censored
observation contributes
`normal_lcdf(mu | upper_limit, observation_sd)`. The latter is the exact Normal
identity
\(\log P(Y \ge U)=\log \Phi((\mu-U)/\sigma)\), expressed in a numerically
stable lower-tail form. It avoids a finite extreme upper-tail probability being
rounded to `log(0)` during initialisation; it does not change the likelihood.
Limits are row-level Stan vectors, so datasets may have no censoring,
one-sided censoring, both sides, or biomarker/observation-specific limits
without branching into separate models.

## Priors

```{r}
library(epikinetics)
priors <- epikinetics_priors()
priors
plot(priors)
```

Population priors are Normal distributions specified by `c(mean, sd)`.
Positive quantities have support above zero, so these Normals are truncated at
zero. Participant SDs and the observation SD have half-Normal priors;
regression coefficients have zero-centred Normal priors.

Stan samples unconstrained standard-Normal raw variables and maps them through
the inverse CDF of these truncated- or half-Normal priors. The implementation
uses complementary probabilities in the upper tail and a value-and-gradient
matched continuation outside the range carrying material Normal probability.
This prevents floating-point rounding from turning leapfrog proposals into
exact zero or infinity. Within the represented prior mass this is the exact
CDF transport and does not alter the scientific model. It makes default
initial values land on the scale of each prior, which matters when a duration
of tens of days and a late-waning rate of thousandths per day coexist in one
model.

This construction separates three ideas that are easy to conflate. Positivity
is the support of the reported quantity; the inverse-CDF map supplies its
truncated-Normal or half-Normal prior; and participant random effects are
non-centred separately through their `z` variables. The standard-Normal raw
variable is the actual Stan parameter and the positive quantity is its
deterministic pushforward, so no additional Jacobian term is required. A
declaration such as `real<lower=0> sigma; sigma ~ normal(0, scale);` has the
same half-Normal distribution (up to its fixed normalising constant), but Stan
then uses its standard exponential lower-bound transform. It does not have the
same unconstrained geometry or scale-aware default initialisation. The exact
transport is retained because that distinction caused the previously observed
frozen-chain behaviour.

Units are explicit:

* baseline is in log2 outcome units;
* time parameters use the input time unit, normally days;
* rates are log2 outcome units per input time unit;
* participant and covariate scales for positive parameters act on log
  parameter ratios.

Override only the scientifically informed quantities while retaining the
complete inspectable object:

```{r}
priors <- epikinetics_priors(
  time_to_peak = c(mean = 12, sd = 4),
  early_waning_rate = c(mean = 0.025, sd = 0.01),
  late_waning_rate = c(mean = 0.003, sd = 0.003),
  observation_sd = 0.75
)
```

For a log2-linear waning rate (r), the response halves after (1/r) time
units. `posterior_parameters()` reports this derived half-life. It is undefined
at exactly zero, but fitted rates are strictly positive; very small rates
appropriately produce very long half-lives.

## Computational implementation

Observations are sorted into contiguous participant ranges. Stan constructs
participant-biomarker kinetics as local model-block values, rather than saving
large transformed-parameter matrices in every posterior draw. `reduce_sum`
partitions the likelihood by participant, preserving locality and amortising
the participant-level parameter construction over each participant's
observations. The model is always compiled with `STAN_THREADS`; users control
parallel chains with `parallel_chains` and within-chain workers with
`threads_per_chain`.

Only the population waning change time is retained as a transformed parameter.
Participant quantities and derived response-scale summaries are reconstructed
from compact posterior draws in R.

The source is installed at `stan/epikinetics.stan` and is deliberately direct:
one curve function, one participant-partitioned likelihood function, and the
hierarchical model.

Every selected participant hierarchy is non-centred. The bundled example has
a median of six observations per participant (some have only three), so a
centred parameterisation would reintroduce the usual group-scale funnel. The
more consequential geometry problem was not centred versus non-centred: it was
asking sparse participant series to identify all six random kinetic effects.
The explicit `participant_parameters` selection avoids that overfitting while
leaving the full model available when study design and diagnostics support it.
