---
title: "Population-level kinetics"
description: >
  Predict conditional population trajectories for biomarkers and covariate
  profiles.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Population-level kinetics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

A population trajectory applies the fitted covariate effects but excludes
participant random effects. It therefore represents the latent kinetics for a
specified population profile, not the arithmetic average of observed people.

```{r, eval=FALSE}
prediction_grid(fit)

population <- predict(
  fit,
  type = "population",
  times = 0:150
)
plot(population)
```

```{r, echo=FALSE, out.width="96%", fig.cap="Conditional population trajectories from an actual package fit. Biomarkers share each panel and covariate profiles define facets. The line is the posterior median latent trajectory; ribbons are pointwise 95% credible intervals and dashed lines are censoring limits."}
doc_figure("documentation-population-kinetics.png")
```

## Prediction profiles

With `newdata = NULL`, `prediction_grid()` keeps observed combinations of
categorical predictors and fixes continuous predictors at participant-level
medians. This avoids silently predicting impossible factor combinations.

```{r, eval=FALSE}
profiles <- data.frame(
  infection_history = c(
    "Infection naive",
    "Previously infected (Pre-Omicron)"
  )
)

population <- predict(
  fit,
  type = "population",
  newdata = profiles,
  times = seq(0, 180, by = 2)
)
```

The stored formula terms, factor levels, contrasts, interactions, and
transformations create the new model matrix. Unknown levels fail early.
`prediction_grid(fit, categorical = "cartesian")` deliberately requests every
fitted-level combination when such extrapolation is scientifically meaningful.

## Summaries and uncertainty

Summarised output contains `mean`, `median`, `lower`, and `upper`, plus
time, biomarker, and original profile columns. The plotting method uses the
median by default; `plot(population, central = "mean")` selects the mean.

The default interval is uncertainty in the latent expected trajectory. Set
`include_observation_noise = TRUE` for a posterior predictive interval for a
future measured value; the returned object and plot subtitle label this
different target explicitly.

Use `summary = FALSE` to retain `.draw` and one trajectory value per posterior
draw. Preserve that identifier when calculating joint quantities over times or
biomarkers.

Response-scale output is the default. It uses log2-spaced axes with natural
response labels, matching the multiplicative outcome scale without compressing
the fitted curves. Every plotting method returns an ordinary `ggplot` object.

Continue with [Individual-level kinetics](individual-kinetics.html) to add
fitted participant effects and observations.
