---
title: "Priors"
description: >
  Inspect, interpret, visualise, and override priors for population kinetics,
  participant variation, covariate effects, and observation error.
output:
  rmarkdown::html_vignette:
    css: epikinetics.css
    toc: true
    toc_depth: 2
lang: en
vignette: >
  %\VignetteIndexEntry{Priors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include=FALSE}
source("vignette-setup.R")
```

## Why inspect the priors?

The six kinetic quantities have different units and typical magnitudes. A
baseline is measured in log2 outcome units, a transition time may span tens of
days, and a late-waning rate may be only a few thousandths per day. Priors
should therefore be inspected on both their parameter scales and the implied
trajectory scale before fitting.

`epikinetics_priors()` returns one complete, printable prior specification:

```{r}
library(epikinetics)

priors <- epikinetics_priors()
priors
```

The object contains three distinct parts:

* population priors for the six biomarker-specific kinetic quantities;
* scale priors for participant effects and regression coefficients; and
* the scale prior for residual observation error.

Passing this object to `prepare_epikinetics_data()` stores the exact prior
specification with the prepared data and subsequently with the fit.

```{r, eval=FALSE}
prepared <- prepare_epikinetics_data(
  data = dat,
  formula = ~ infection_history,
  lower_limit = 5,
  upper_limit = 2560,
  priors = priors
)
```

## Population priors

Population priors are Normal distributions specified as `c(mean, sd)`.
Baseline is unconstrained. The other five population quantities are positive,
so their Normal distributions are truncated at zero:

| Quantity | Units and interpretation |
|---|---|
| `baseline` | Expected log2 outcome at the focal exposure. |
| `time_to_peak` | Time from exposure to peak, in the input time unit. |
| `waning_duration` | Time from peak to the early/late waning switch. |
| `boost_rate` | Log2 increase per input time unit before the peak. |
| `early_waning_rate` | Positive magnitude of the first log2 decline rate. |
| `late_waning_rate` | Positive magnitude of the later log2 decline rate. |

Every biomarker has its own population parameter, with the configured prior
applied to each. The priors are independent before conditioning on data.

For a log2-linear waning rate (r), the response halves after (1/r) time units.
This can help translate a rate prior into a scientifically familiar half-life.
Very small positive rates appropriately imply very long half-lives.

## Prior trajectory check

The plot method draws population kinetic parameters and passes every draw
through the piecewise kinetic curve. It deliberately excludes participant
effects and residual observation error. The result is therefore a prior check
for the latent population trajectory, not a full prior predictive distribution
for future measurements.

```{r prior-trajectories, fig.cap="Latent population trajectories implied by the default kinetic-parameter priors. The line is the pointwise median and the ribbon is the pointwise 95% prior interval; participant variation and observation error are excluded."}
set.seed(2026)
plot(
  priors,
  scale = "response",
  reference_value = 1
)
```

On the response scale, the vertical axis uses log2 spacing while retaining
natural measurement labels. This is important: uncertainty that is additive
on the fitted log2 scale becomes multiplicative after transforming back to the
response scale. A linear response axis can make a coherent prior look like a
nearly flat median under an enormous ribbon.

Increasing `ndraws` makes the estimated quantiles more stable; it does not make
the prior distribution narrower. Use `probs` to request another clearly
labelled interval when that is useful, or use `scale = "model"` to inspect the
same trajectories directly in log2 units:

```{r, eval=FALSE}
plot(priors, ndraws = 2000, probs = c(0.05, 0.95))
plot(priors, scale = "model")
```

If the population trajectory interval is scientifically implausible, revise
the underlying parameter priors rather than clipping the plot. Participant and
observation priors should be assessed separately because they answer different
questions and can otherwise obscure which component is driving the variation.

## Participant and covariate scales

Participant standard deviations have half-Normal priors. They act additively
for baseline and on the log scale for positive time and rate quantities. For a
positive quantity, a participant effect of (u) therefore multiplies the
conditional population value by \(\exp(u)\).

The default participant hierarchy is selected during data preparation. It
includes baseline, boost rate, early-waning rate, and late-waning rate; peak
timing and waning duration are shared across participants by default. Priors
for inactive participant effects are retained in the inspectable prior object
but do not create parameters in Stan.

Regression coefficients have zero-centred Normal priors. A baseline coefficient
is an additive shift in log2 outcome units. Coefficients for positive timing or
rate quantities are log ratios, so exponentiating a coefficient gives its
multiplicative effect. `covariate_sd` specifies the prior standard deviation for
each kinetic quantity on the corresponding link scale.

## Observation error

`observation_sd` is the half-Normal scale for residual Normal error on the
model's log2 scale. It is not included in the population trajectory plot above.
Including it would answer a different question: the distribution of a future
observed measurement around the latent curve.

After fitting, `predict(..., include_observation_noise = FALSE)` returns latent
trajectory uncertainty, while `include_observation_noise = TRUE` returns an
explicitly labelled posterior predictive interval. The same distinction should
be maintained when reasoning about priors.

## Overriding defaults

Override only quantities for which there is a scientific or measurement-based
reason. Unspecified fields retain their defaults:

```{r}
priors <- epikinetics_priors(
  time_to_peak = c(mean = 12, sd = 4),
  early_waning_rate = c(mean = 0.025, sd = 0.01),
  late_waning_rate = c(mean = 0.003, sd = 0.003),
  participant_sd = c(
    baseline = 0.75,
    time_to_peak = 0.35,
    waning_duration = 0.50,
    boost_rate = 0.35,
    early_waning_rate = 0.50,
    late_waning_rate = 0.75
  ),
  covariate_sd = c(
    baseline = 0.50,
    time_to_peak = 0.35,
    waning_duration = 0.50,
    boost_rate = 0.35,
    early_waning_rate = 0.50,
    late_waning_rate = 0.75
  ),
  observation_sd = 0.75
)
priors
```

Named participant and covariate scale vectors must contain all six quantities.
This makes omissions visible rather than silently applying a value to the wrong
parameter.

## Stan parameterisation

Stan samples unconstrained standard-Normal raw variables and maps them through
the inverse CDF of the configured truncated- or half-Normal prior. The
implementation uses complementary probabilities in the upper tail and a
value-and-gradient matched continuation beyond the range carrying material
Normal probability. Within that range, it is the exact inverse-CDF transport.

This separates three concepts: positivity defines the support of a reported
quantity; the inverse-CDF map supplies its prior; and participant random effects
are non-centred separately through their standardised effects. The transport
also puts Stan's default initial values on the scale of each configured prior,
which avoids poor geometry when timing, rate, and scale parameters differ by
orders of magnitude.

The [Kinetics model and statistical structure](model.html) vignette describes
how these parameters enter the hierarchical model and likelihood.
