---
title: "SARS-CoV-2 antibody model: replicating paper figures"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{SARS-CoV-2 antibody model: replicating paper figures}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Fitting the model

To initialise a model object, the only required argument is a path to the data in CSV format, or a [data.table]("https://cran.r-project.org/web/packages/data.table/index.html"). See [scova](../reference/scova.html) for all available arguments. In this vignette we use a dataset representing the Delta wave which is installed with this package, specifying a hierarchical model that just looks at the effect of infection history.

The `fit` method then has the same function signature as the underlying [cmdstanr::sample](https://mc-stan.org/cmdstanr/reference/model-method-sample.html) method. Here we specify a relatively small number of iterations of the algorithm to limit the time it takes to compile this vignette.

```{r, warning=FALSE, message=FALSE, results="hide"}
dat <- data.table::fread(system.file("delta_full.rds", package = "epikinetics"))
mod <- epikinetics::scova$new(data = dat, covariate_formula = ~0 + infection_history)
delta <- mod$fit(parallel_chains = 4,
                 iter_warmup = 50,
                 iter_sampling = 200,
                 threads_per_chain = 4)
```

# Figure 2
Figure 2 from the paper shows population level fits for each wave, disaggregated by infection history
and titre type. Here we reproduce the facets for the Delta wave. Once the model has been fitted, simulate population trajectories
using the fitted population parameters.

```{r}
res <- mod$simulate_population_trajectories()
head(res)
```

See [data](data.html) for the definition of the returned columns.

Using `ggplot2`:


```{r}
library(ggplot2)
custom_theme <- theme_linedraw() + theme(
    legend.position = "bottom",
    text = element_text(size = 8, family = "Helvetica"),
    strip.text.x.top = element_text(size = 8, family = "Helvetica"),
    strip.text.x = element_text(size = 8, family = "Helvetica"),
    strip.background = element_rect(fill = "white"),
    strip.text = element_text(colour = 'black'),
    strip.placement = "outside",
    legend.box = "vertical",
    legend.margin = margin(),
)

custom_palette <- c("#CC6677", "#DDCC77", "#88CCEE")

plot_data <- res
plot_data[, titre_type := forcats::fct_relevel(
  titre_type,
  c("Ancestral", "Alpha", "Delta"))]

ggplot(data = plot_data) +
  geom_line(aes(x = t,
                y = me,
                colour = titre_type)) +
  geom_ribbon(aes(x = t,
                  ymin = lo,
                  ymax = hi,
                  fill = titre_type), alpha = 0.65) +
  scale_y_continuous(
    trans = "log2",
    breaks = c(40, 80, 160, 320, 640, 1280,
               2560, 5120),
    labels = c("40", "80", "160", "320", "640", "1280", "2560", "5120"),
    limits = c(40, 10240)) +
  scale_x_continuous(breaks = c(0, 30, 60, 90, 120),
                     labels = c("0", "30", "60", "90", "120"),
                     expand = c(0, 0)) +
  coord_cartesian(clip = "off") +
  labs(x = "Time since last exposure (days)",
       y = expression(paste("Titre (IC"[50], ")"))) +
  facet_wrap(infection_history ~ titre_type) +
  scale_colour_manual(values = custom_palette) +
  scale_fill_manual(values = custom_palette) +
  guides(colour = "none", fill = "none") +
  custom_theme
```

# Figure 4

Figure 4 shows the relationship between infection history and titre values at peak and set points, for different
titre types. We first simulate estimates for these from the fitted model:

```{r}
res <- mod$population_stationary_points()
head(res)
```

The values we're going to plot are the mean peak titre values (`mu_p_me`) and mean set point titre values (`mu_s_me`), for different titre types and infection histories. See [data](data.html) for a full definition of all the returned columns. Using `ggplot2`:

```{r}
plot_data <- res[, titre_type := forcats::fct_relevel(
  titre_type,
  c("Ancestral", "Alpha", "Delta"))]

ggplot(data = plot_data, aes(
  x = mu_p, y = mu_s,
  colour = titre_type)) +
  geom_density_2d(
    aes(
      group = interaction(
        infection_history,
        titre_type))) +
  geom_point(data = plot_data[.draw <= 2000],
             alpha = 0.05, size = 0.2) +
  geom_point(aes(x = mu_p_me, y = mu_s_me,
                 shape = infection_history),
             colour = "black") +
  geom_path(aes(x = mu_p_me, y = mu_s_me,
                group = titre_type),
            colour = "black") +
  geom_vline(xintercept = 2560, linetype = "twodash", colour = "gray30") +
  scale_x_continuous(
    trans = "log2",
    breaks = c(40, 80, 160, 320, 640, 1280, 2560, 5120, 10240),
    labels = c(expression(" " <= 40),
               "80", "160", "320", "640", "1280", "2560", "5120", "10240"),
    limits = c(NA, 10240)) +
  geom_hline(yintercept = 2560, linetype = "twodash", colour = "gray30") +
  scale_y_continuous(
    trans = "log2",
    breaks = c(40, 80, 160, 320, 640, 1280, 2560, 5120, 10240),
    labels = c(expression(" " <= 40),
               "80", "160", "320", "640", "1280", "2560", "5120", "10240"),
    limits = c(NA, 5120)) +
  custom_theme +
  scale_shape_manual(values = c(1, 2, 3)) +
  labs(x = expression(paste("Population-level titre value at peak (IC"[50], ")")),
       y = expression(paste("Population-level titre value at set-point (IC"[50], ")"))) +
  scale_colour_manual(values = custom_palette) +
  guides(colour = guide_legend(title = "Titre type", override.aes = list(alpha = 1, size = 1)),
         shape = guide_legend(title = "Infection history"))
```

# Figure 5
Figure 5 uses simulations of individual trajectories to derive population level titre values. This can take a while so here we set the maximum number of draws to just 250 for speed.

```{r}
res <- mod$simulate_individual_trajectories(n_draws = 250)
head(res)
```

See [data](data.html) for a definition of all the returned columns.
Figure 5 A plots the derived population trajectories. Here we replicate a portion of the graph from the paper,
from the minimum date in the dataset until the date at which the next variant emerged.

```{r}
date_delta <- lubridate::ymd("2021-05-07")
date_ba2 <- lubridate::ymd("2022-01-24")

res$wave <- "Delta"
dat$wave <- "Delta"
plot_data <- merge(
  res, dat[, .(
    min_date = min(date), max_date = max(date)), by = wave])[
  , .SD[calendar_date >= min_date & calendar_date <= date_ba2], by = wave]

plot_data[, titre_type := forcats::fct_relevel(
  titre_type,
  c("Ancestral", "Alpha", "Delta"))]

ggplot() + geom_line(
  data = plot_data,
  aes(x = calendar_date,
      y = me,
      group = interaction(titre_type, wave),
      colour = titre_type),
  alpha = 0.2) +
  geom_ribbon(
    data = plot_data,
    aes(x = calendar_date,
        ymin = lo,
        ymax = hi,
        group = interaction(titre_type, wave)
    ),
    alpha = 0.2) +
  labs(title = "Population-level titre values",
       tag = "A",
       x = "Date",
       y = expression(paste("Titre (IC"[50], ")"))) +
  scale_colour_manual(
    values = custom_palette) +
  scale_fill_manual(
    values = custom_palette) +
  scale_x_date(
    date_labels = "%b %Y",
    limits = c(min(dat$date), date_ba2)) +
  geom_smooth(
    data = plot_data,
    aes(x = calendar_date,
        y = me,
        fill = titre_type,
        colour = titre_type,
        group = interaction(titre_type, wave)),
    alpha = 0.5, span = 0.2) +
  custom_theme +
  guides(colour = guide_legend(title = "Titre type"), fill = "none",)
```

Figure 5 C shows counterfactual trajectories varying vaccination timings. To do this, we simulate
individual trajectories with different `time_shift` arguments. Here we set the maximum number of draws to just 50 for speed.

```{r}
time_shift_values <- seq(-75, 75, by = 15)
indices <- seq_along(time_shift_values)

results_list <- lapply(indices, function(index) {
  shift <- time_shift_values[index]
  result <- mod$simulate_individual_trajectories(n_draws = 50, time_shift = shift)
  result[, time_shift:= shift]
  result
})

combined_data <- data.table::data.table(data.table::rbindlist(results_list))
```

Plotting the median values:

```{r}
plot_data <- combined_data[calendar_date == date_delta]
plot_data <- plot_data[, titre_type := forcats::fct_relevel(
    titre_type,
    c("Ancestral", "Alpha", "Delta"))]

ggplot(
  data = plot_data) +
  geom_line(aes(
    x = time_shift, y = me, colour = titre_type)) +
  geom_pointrange(aes(
    x = time_shift, y = me, ymin = lo,
    ymax = hi, colour = titre_type)) +
  scale_x_continuous(
    breaks = time_shift_values) +
  scale_y_continuous(
    trans = "log2",
    breaks = c(40, 80, 160, 320, 640, 1280),
    labels = c("40", "80", "160", "320", "640", "1280")) +
  scale_colour_manual(
    values = custom_palette) +
  labs(title = "Population titre values varying vaccine timings",
       x = "Time shift (days)",
       y = expression(paste("Titre (IC"[50], ")"))) +
  custom_theme +
  theme(legend.position = "none") +
  guides(colour = guide_legend(title = "Titre Type", nrow = 1, override.aes = list(alpha = 1, size = 1))) +
  labs(tag = "C",
       title = "Counterfactual population-level titres varying vaccination timings")
```
